---
title: "Get started with CraftGRN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with CraftGRN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

CraftGRN is an R package for constructing and comparing condition-specific gene
regulatory networks from matched ATAC-seq and RNA-seq data. The workflow follows
three modules:

1. Predict TF binding sites from footprint or peak signal.
2. Connect TF binding sites to candidate target genes.
3. Learn regulatory topics and visualize differential GRNs.

The full pipeline is designed around a project YAML file that records paths,
sample metadata, threshold values, genome settings, and output directories.

```{r install, eval=FALSE, purl=FALSE}
remotes::install_github("oncologylab/craftgrn")
library(craftgrn)
```

## Demo data

CraftGRN keeps demo datasets outside the source package so installation remains
small and CRAN-friendly. The package helper reports any configured external
demo bundles:

```{r demo-data}
craftgrn_demo_data_info()
```

No external demo bundle is currently configured. To run your own project, point
CraftGRN at a project-level YAML file and use the normal Module 1 preparation
function:

```{r demo-data-load}
config <- "project.yaml"

omics <- load_prep_multiomic_data(
  config = config,
  label_col = "strict_match_rna",
  do_preprocess = FALSE,
  verbose = TRUE
)
```

To run Module 1:

```{r demo-data-module1}
module1 <- predict_tfbs(
  omics_data = omics,
  out_dir = "predict_tf_binding_sites",
  output_format = "auto",
  write_stats = FALSE,
  verbose = TRUE
)
```

Troubleshooting demo data:

- If `craftgrn_demo_data_info()` returns zero rows, no public demo bundle is
  currently advertised by this package version.
- If file paths fail after moving the project, pass the `project.yaml` path
  explicitly. A portable config should use `base_dir: "."` so paths are
  relative to the config location.
- If the computer has limited memory, first run only `load_prep_multiomic_data()`
  and Module 1. Module 2 is larger because it computes TF-target correlations
  before restricting FP-target candidates to predicted TFBS.

## Module 1: Predict TF binding sites

Module 1 combines normalized ATAC signal, RNA expression, footprint summaries,
and sample metadata into a multiomic object. It aligns footprint sites, builds
condition-specific binding flags, normalizes footprint scores, and predicts TF
binding sites with a single user-facing workflow. Internally, CraftGRN first
correlates each aligned footprint with the TFs represented by its canonical
motifs, applies configured correlation cutoffs, and labels footprints with at
least one canonical-bound TF. By default, only these canonical-bound footprints
are used for the all-expressed-TF prediction step.

```{r module-1-figure, echo = FALSE, out.width = "100%", eval = TRUE, fig.alt = "Module 1 workflow for preparing multiomic data, processing footprint evidence, and predicting TF binding sites."}
knitr::include_graphics("https://raw.githubusercontent.com/oncologylab/craftgrn/main/figures/module_1.svg")
```

### Inputs

Module 1 expects:

- A merged ATAC peak master table with normalized signal and overlap flags.
- A normalized RNA-seq count or expression matrix.
- A directory of per-condition motif footprint score and binding overview files
  generated by TOBIAS or a TOBIAS-compatible workflow.
- A metadata table mapping samples to conditions.
- A YAML configuration file with file paths, thresholds, and genome settings.
  The TFBS workflow uses `threshold_fp_tf_corr_r`, and can also use optional
  `threshold_fp_tf_corr_p` and `threshold_fp_tf_corr_fdr` fields when a project
  should require p-value or adjusted p-value support.

Supported built-in motif resources include `hg38` and `mm10`.

For large projects, keep `write_stats = FALSE` and use `output_format = "auto"`.
Module 1 keeps full correlation tables as audit outputs, but the primary
handoff to Module 2 is `predicted_tfbs`: a compact table with only aligned
canonical-bound FPs and the TF names predicted to bind each FP.

### Run Module 1

```{r module-1}
data_object <- load_prep_multiomic_data(
  config = "craftgrn_grn.yaml",
  genome = "hg38",
  gene_symbol_col = "HGNC"
)

module1 <- predict_tfbs(
  data_object,
  out_dir = "predict_tf_binding_sites",
  output_format = "auto"
)

module1_qc <- build_module1_qc_report(
  module1,
  output_dir = file.path("predict_tf_binding_sites", "reports"),
  scan_predicted_tfbs = TRUE
)
```

### Outputs

Module 1 creates:

- A compact multiomic object with normalized signal matrices, binary flags, and
  feature annotations.
- Compact aligned footprint site, score, binding, and motif-support tables.
- Canonical-bound footprint summaries, including all canonical motifs retained
  for each aligned footprint.
- `predicted_tfbs`, the compact FP-to-TF handoff used by Module 2.
- Optional full TFBS correlation statistics for audit and debugging.
- `module1_qc_report.html`: an HTML QC report summarizing run parameters, input gates, motif-supported canonical support, correlation diagnostics, chunked predicted TFBS integrity, top TFs/FPs, condition support, motif complexity, warning checks, and related artifacts. The report uses static processing funnels, density curves, scatter summaries, heatmaps, lollipop rank plots, and cumulative curves.
- QC summaries reporting how many FPs, motif-supported pairs, canonical-bound
  FPs, prediction pairs, and predicted TFBS rows pass each step.

## Module 2: Connect TFs to target genes

Module 2 treats TF-target prediction as a compact relational problem. It joins
the Module 1 `predicted_tfbs` relation with TF-target expression correlations,
restricted FP-target correlations, TSS-window evidence, and optional generic
regulatory-prior evidence. GeneHancer, HiC, promoter-capture, CRISPR enhancer
screens, or user tables are all represented as regulatory priors rather than
separate modes.

```{r module-2-figure, echo = FALSE, out.width = "100%", eval = TRUE, fig.alt = "Module 2 workflow for linking predicted TF binding sites to target genes using expression, genomic distance, and optional regulatory prior evidence."}
knitr::include_graphics("https://raw.githubusercontent.com/oncologylab/craftgrn/main/figures/module_2.svg")
```

### Inputs

Module 2 uses:

- The compact multiomic object generated by Module 1.
- `predicted_tfbs`, with one row per predicted FP-TF binding event.
- A target gene TSS annotation table.
- Optional generic regulatory-prior FP-target links.
- YAML thresholds for TF-target and FP-target correlation filtering.

### Process

Module 2 first computes TF expression to target expression correlations and
filters TF-target pairs. Only after that filter does it build FP-target
candidates from FPs bound by the surviving TFs. FP score to target expression
correlations are computed only for those restricted FP-target candidates within
the TSS window or supported by regulatory prior evidence.

```{r module-2, eval = FALSE}
module2 <- link_tf_targets(
  multiomic_data = module1$omics_data,
  predicted_tfbs = module1$predicted_tfbs,
  gene_tss = gene_tss,
  regulatory_prior = regulatory_prior,
  project_config = "project.yaml",
  output_dir = "connect_tf_target_genes"
)

hnf4a_links <- query_module2_links(module2, tf = "HNF4A")

module2_qc <- build_module2_qc_report(
  module2,
  multiomic_data = module1$omics_data,
  output_dir = file.path("connect_tf_target_genes", "reports"),
  scan_large_tables = TRUE,
  validate_integrity = TRUE
)
```

### Outputs

Module 2 produces normalized compact tables:

- `module2_tf_target_corr`: one row per TF-target correlation.
- `module2_fp_target_candidates`: one row per FP-target candidate with
  distance-to-TSS and prior evidence.
- `module2_fp_target_corr`: one row per FP-target correlation.
- `module2_links`: final TF-FP-target links as compact keys and pass flags.
- `module2_condition_activity`: long condition-level activity flags.
- `module2_qc_summary`: counts before and after each filter.
- `module2_qc_report.html`: an HTML QC report summarizing compact handoff checks, TF-target and FP-target correlation filters, FP-target candidate source and distance-to-TSS evidence, final-link integrity checks, condition activity, warning checks, top TF/target/FP summaries, output manifests, and related HTML browser reports. The report uses relational flow diagrams, density and cumulative distance plots, scatter summaries, heatmaps, and lollipop rank plots.

## Module 3: Learn regulatory topics and visualize differential GRNs

Module 3 compares regulatory links between conditions, builds joint RNA and
footprint document-term matrices, trains topic models, assigns links to topics,
and summarizes topic-specific pathway and master TF programs.

```{r module-3-figure, echo = FALSE, out.width = "100%", eval = TRUE, fig.alt = "Module 3 workflow for comparing regulatory links, training topic models, and summarizing differential GRN programs."}
knitr::include_graphics("https://raw.githubusercontent.com/oncologylab/craftgrn/main/figures/module_3.svg")
```

### Inputs

Module 3 expects:

- Per-condition TF->TFBS->target link matrices from Module 2.
- A comparison metadata table defining pairwise contrasts.
- TF motif cluster assignments from Module 1.

### Train topic models

The topic modeling workflow builds documents such as
`comparison x TF-cluster x direction`. Terms are genes and TFBS features, and
weights are derived from RNA log2 fold changes and footprint or peak signal
deltas. The model balances RNA and footprint contributions before converting
weights into pseudocounts for topic modeling.
By default, WarpLDA uses CraftGRN's native `warp_omp` sampler, which keeps
doc/word Metropolis-Hastings semantics while using OpenMP threads inside each
model fit. For fixed-seed validation runs, set
`warplda_sampler = "warp_ref"`; this native sequential reference sampler is
much slower than `warp_omp`.

```{r module-3-topic-models}
train_topic_models(
  Kgrid = c(2:10, 15, 20, 25, 30),
  input_dir = "differential_links",
  output_dir = "topic_models"
)

extract_regulatory_topics(
  k = 10,
  link_dir = "differential_links",
  model_dir = "topic_models",
  output_dir = "final_topics",
  qc = TRUE
)
```

### Summarize pathways and master TFs

```{r module-3-summaries}
run_diff_grn_pathway_analysis(
  link_dir = "differential_links",
  output_dir = "pathway_analysis"
)

run_diff_grn_master_tf_summary(
  link_dir = "differential_links",
  output_dir = "master_tf_summary"
)
```

### Outputs

Module 3 creates:

- Differential link tables and comparison-level plots.
- Model selection metrics across topic-number choices.
- Topic-term and topic-link probability tables.
- Topic assignments for regulatory links.
- Pathway enrichment summaries and plots.
- Master TF connectivity summaries and plots.
- Topic-stratified differential GRN visualizations.

## Next steps

After running the three modules, inspect the model selection metrics, choose a
topic number, review topic-specific pathway enrichments, and use master TF
summaries to prioritize condition-specific regulatory programs for follow-up.
