---
title: "Get started with CraftGRN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with CraftGRN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

CraftGRN is an R package for constructing and comparing condition-specific gene
regulatory networks from matched ATAC-seq and RNA-seq data. The workflow follows
three modules:

1. Predict TF binding sites from footprint or peak signal.
2. Connect TF binding sites to candidate target genes.
3. Learn regulatory topics and visualize differential GRNs.

The full pipeline is designed around a project YAML file that records paths,
sample metadata, threshold values, genome settings, and output directories.

```{r install}
remotes::install_github("oncologylab/craftgrn")
library(craftgrn)
```

## Module 1: Predict TF binding sites

Module 1 combines normalized ATAC signal, RNA expression, footprint summaries,
and sample metadata into a multiomic object. It then collapses redundant motif
footprints, builds condition-specific binding flags, normalizes footprint
scores, and correlates TF expression with footprint or peak signal across
matched conditions.

```{r module-1-figure, echo = FALSE, out.width = "100%", eval = TRUE}
knitr::include_graphics("../figures/module_1.svg")
```

### Inputs

Module 1 expects:

- A merged ATAC peak master table with normalized signal and overlap flags.
- A normalized RNA-seq count or expression matrix.
- A directory of per-condition motif footprint score and binding overview files
  generated by TOBIAS or a TOBIAS-compatible workflow.
- A metadata table mapping samples to conditions.
- A YAML configuration file with file paths, thresholds, and genome settings.

Supported built-in motif resources include `hg38` and `mm10`.

### Run Module 1

```{r module-1}
data_object <- load_prep_multiomic_data(
  config = "craftgrn_grn.yaml",
  genome = "hg38",
  gene_symbol_col = "HGNC"
)

data_object <- correlate_tf_to_fp(
  data_object,
  qc = TRUE
)
```

### Outputs

Module 1 creates:

- A combined multiomic object with normalized signal matrices, binary flags, and
  feature annotations.
- Collapsed raw footprint score matrices and footprint merging summaries.
- Collapsed, filtered, and normalized footprint score matrices.
- Footprint binding flag matrices and footprint annotation tables.
- TF motif cluster assignments.
- Gene expression flags.
- Per-TF binding probability summaries and QC plots.

## Module 2: Connect TFs to target genes

Module 2 links predicted TF binding sites to candidate target genes. Candidate
links can come from GeneHancer enhancer-gene mappings, a genomic TSS window, or
user-supplied 3D chromatin interaction loops. Links are retained when they pass
expression, binding, footprint or peak signal, and correlation filters.

```{r module-2-figure, echo = FALSE, out.width = "100%", eval = TRUE}
knitr::include_graphics("../figures/module_2.svg")
```

### Inputs

Module 2 uses:

- The multiomic object generated by Module 1.
- Candidate enhancer-gene, TSS-window, or chromatin-loop priors.
- YAML thresholds for TF expression, target expression, peak overlap, footprint
  signal, peak signal, and correlation filtering.

### Process

For each candidate TF->TFBS->target link, CraftGRN evaluates whether the TF and
target gene are expressed in each condition, whether the TFBS is bound, whether
footprint or peak signal passes configured cutoffs, and whether TF-target and
TFBS-target signal relationships are supported across matched conditions.

### Outputs

Module 2 produces:

- A TF->TFBS->target overview table with condition-specific link status flags.
- Correlation statistics for retained links.
- Active-link count summaries by TF and condition.
- Per-condition link matrices for visualization and differential analysis.

## Module 3: Learn regulatory topics and visualize differential GRNs

Module 3 compares regulatory links between conditions, builds joint RNA and
footprint document-term matrices, trains topic models, assigns links to topics,
and summarizes topic-specific pathway and master TF programs.

```{r module-3-figure, echo = FALSE, out.width = "100%", eval = TRUE}
knitr::include_graphics("../figures/module_3.svg")
```

### Inputs

Module 3 expects:

- Per-condition TF->TFBS->target link matrices from Module 2.
- A comparison metadata table defining pairwise contrasts.
- TF motif cluster assignments from Module 1.

### Train topic models

The topic modeling workflow builds documents such as
`comparison x TF-cluster x direction`. Terms are genes and TFBS features, and
weights are derived from RNA log2 fold changes and footprint or peak signal
deltas. The model balances RNA and footprint contributions before converting
weights into pseudocounts for topic modeling.

```{r module-3-topic-models}
train_topic_models(
  Kgrid = c(2:10, 15, 20, 25, 30),
  input_dir = "differential_links",
  output_dir = "topic_models"
)

extract_regulatory_topics(
  k = 10,
  link_dir = "differential_links",
  model_dir = "topic_models",
  output_dir = "final_topics",
  qc = TRUE
)
```

### Summarize pathways and master TFs

```{r module-3-summaries}
run_diff_grn_pathway_analysis(
  link_dir = "differential_links",
  output_dir = "pathway_analysis"
)

run_diff_grn_master_tf_summary(
  link_dir = "differential_links",
  output_dir = "master_tf_summary"
)
```

### Outputs

Module 3 creates:

- Differential link tables and comparison-level plots.
- Model selection metrics across topic-number choices.
- Topic-term and topic-link probability tables.
- Topic assignments for regulatory links.
- Pathway enrichment summaries and plots.
- Master TF connectivity summaries and plots.
- Topic-stratified differential GRN visualizations.

## Next steps

After running the three modules, inspect the model selection metrics, choose a
topic number, review topic-specific pathway enrichments, and use master TF
summaries to prioritize condition-specific regulatory programs for follow-up.
